---
title: "Shohei Ohtani's 2024 Pitching What-If
author: "Marco Paul Perez"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load Necessary Libraries
```{r}
library(dplyr)
library(ggplot2)
library(tidyr)
```

Set working directory (adjust the path accordingly)
```{r}
setwd('/Users/paulperezgarcia/Documents/Data Playbook/Ohtani Pitching')
```

Import pitching statistics from CSV files
```{r}
stats_2021 <- read.csv('2021 Pitching Stats.csv')
stats_2022 <- read.csv('2022 Pitching Stats.csv')
stats_2023 <- read.csv('2023 Pitching Stats.csv')
stats_2024 <- read.csv('2024 Pitching Stats.csv')
```

Select only important columns for analysis
```{r}
selected_columns <- c("Player", "W", "ERA", "IP", "WHIP", "FIP")

stats_2021 <- stats_2021[, selected_columns]
stats_2022 <- stats_2022[, selected_columns]
stats_2023 <- stats_2023[, selected_columns]
stats_2024 <- stats_2024[, selected_columns]
```

Remove duplicate player entries (if they played for multiple teams in a season)
```{r}
stats_2021 <- stats_2021 %>%
  distinct(Player, .keep_all = TRUE)
stats_2022 <- stats_2022 %>%
  distinct(Player, .keep_all = TRUE)
stats_2023 <- stats_2023 %>%
  distinct(Player, .keep_all = TRUE)
stats_2024 <- stats_2024 %>%
  distinct(Player, .keep_all = TRUE) %>%
  filter(IP >= 100)
```

Identify players who played in all four seasons
```{r}
common_players <- Reduce(intersect, list(stats_2021$Player, stats_2022$Player, stats_2023$Player, stats_2024$Player))
```

Filter the 2024 stats to include only common players
```{r}
stats_2024_common <- stats_2024 %>%
  filter(Player %in% common_players)
```

Standardize key performance variables
```{r}
stats_2024_z <- stats_2024_common %>%
  mutate(
    ERA_z = -scale(ERA),
    WHIP_z = -scale(WHIP),
    W_z = scale(W),
    IP_z = scale(IP),
    FIP_z = -scale(FIP),
    Weighted_Score = ERA_z * .35 +
                     WHIP_z * .20 +
                     W_z * .05 +
                     IP_z * .15 +
                     FIP_z * .25
  )
```

Results from the Weighted_Score analysis
```{r}
print(stats_2024_z %>%
  dplyr::select(Player, Weighted_Score) %>%
  arrange(desc(Weighted_Score)))
```

Create a dataset for each season that only contains the common_players
```{r}
stats_2021_common <- stats_2021[stats_2021$Player %in% common_players, ]
stats_2022_common <- stats_2022[stats_2022$Player %in% common_players, ]
stats_2023_common <- stats_2023[stats_2023$Player %in% common_players, ]
```

Merge yearly stats into a single dataset
```{r}
merged_stats <- merge(stats_2021_common, stats_2022_common, by="Player", suffixes = c("_2021", "_2022"))
merged_stats <- merge(merged_stats, stats_2023_common, by="Player")
```

Compute weighted averages for ERA, FIP, and WHIP. Average innings pitched (IP) and wins (W)
```{r}
merged_stats$ERA_weighted <- merged_stats$ERA_2021 * 0.2 + merged_stats$ERA_2022 * 0.3 + merged_stats$ERA * 0.5
merged_stats$FIP_weighted <- merged_stats$FIP_2021 * 0.2 + merged_stats$FIP_2022 * 0.3 + merged_stats$FIP * 0.5
merged_stats$WHIP_weighted <- merged_stats$WHIP_2021 * 0.2 + merged_stats$WHIP_2022 * 0.3 + merged_stats$WHIP * 0.5
merged_stats$IP_avg <- rowMeans(merged_stats[, c("IP_2021", "IP_2022", "IP")])
merged_stats$W_avg <- rowMeans(merged_stats[, c("W_2021", "W_2022", "W")])
```

Select final statistics
```{r}
final_stats <- merged_stats[, c("Player", "ERA_weighted", "FIP_weighted", "WHIP_weighted", "IP_avg", "W_avg")]
```

Merge with 2024 weighted scores and Rename Column
```{r}
final_stats <- merge(final_stats, stats_2024_z[, c("Player", "Weighted_Score")], by = "Player", all.x = TRUE)

final_stats <- final_stats %>%
  rename(Weighted_Score2024 = Weighted_Score)
```

Develop a multiple regression model to predict 2024 weighted scores
```{r}
regression_model <- lm(Weighted_Score2024 ~ ERA_weighted + FIP_weighted + WHIP_weighted + IP_avg + W_avg, data = final_stats)

summary(regression_model)
```

Predict weighted score for Shohei Ohtani
```{r}
player_name <- "Shohei Ohtani"

ohtani_2021 <- stats_2021[stats_2021$Player == player_name, ]
ohtani_2022 <- stats_2022[stats_2022$Player == player_name, ]
ohtani_2023 <- stats_2023[stats_2023$Player == player_name, ]

ohtani_weighted_stats <- data.frame(
  ERA_weighted = (ohtani_2021$ERA * 0.2) + (ohtani_2022$ERA * 0.3) + (ohtani_2023$ERA * 0.5),
  FIP_weighted = (ohtani_2021$FIP * 0.2) + (ohtani_2022$FIP * 0.3) + (ohtani_2023$FIP * 0.5),
  WHIP_weighted = (ohtani_2021$WHIP * 0.2) + (ohtani_2022$WHIP * 0.3) + (ohtani_2023$WHIP * 0.5),
  IP_avg = mean(c(ohtani_2021$IP, ohtani_2022$IP, ohtani_2023$IP)),
  W_avg = mean(c(ohtani_2021$W, ohtani_2022$W, ohtani_2023$W))
)

ohtani_prediction <- predict(regression_model, ohtani_weighted_stats)

print(paste("Predicted Weighted_Score2024 for Shohei Ohtani:", round(ohtani_prediction, 4)))
```

Add Shohei Ohtani to data frame with the rest of the players
```{r}
ohtani_weighted_stats$Player <- "Shohei Ohtani"

ohtani_weighted_stats$Weighted_Score2024 <- ohtani_prediction

final_stats_extended <- rbind(final_stats, ohtani_weighted_stats)
```

Prediction vs Actual Stats
```{r}
final_stats_prediction_comp <- final_stats %>%
  mutate(Weighted_Score_Prediction = predict(regression_model, newdata = final_stats))
```

```{r}
ggplot(final_stats_prediction_comp, aes(x = Weighted_Score2024, y = Weighted_Score_Prediction)) +
  geom_point(color = "dodgerblue", alpha = 0.7) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "black", linewidth = 1) + 
  labs(
    title = "Actual vs. Predicted Weighted Scores (2024)",
    x = "Actual Weighted Score",
    y = "Predicted Weighted Score"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5)
  )
```

Create a Data Frame of how would it look vs Zach Wheeler (highest rated pitcher) and Total Pitchers 
```{r}
total_mean <- mean(final_stats_extended$Weighted_Score2024)
ohtani_score <- final_stats_extended[final_stats_extended$Player == "Shohei Ohtani", ]$Weighted_Score2024
wheeler_score <- final_stats_extended[final_stats_extended$Player == "Zack Wheeler", ]$Weighted_Score2024

comparison_df <- data.frame(
  Category = c("Zack Wheeler", "Shohei Ohtani", "Total Mean of Pitchers"),
  Score = c(wheeler_score, ohtani_score, total_mean))
```

Bar Plot comparison between Ohtani, Wheeler and the total players average 
```{r}
comparison_df$Category <- factor(comparison_df$Category, levels = comparison_df$Category[order(comparison_df$Score, decreasing = TRUE)])


baseball_colors <- c("Shohei Ohtani" = "dodgerblue",
                     "Zack Wheeler" = "#FFD700", 
                     "Total Mean of Pitchers" = "#BA0021")

ggplot(comparison_df, aes(x = Score, y = Category, fill = Category)) +
  geom_col(show.legend = FALSE) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "black", linewidth = 1) +
  geom_text(aes(label = round(Score, 2)), hjust = -.05, 
            size = 3, fontface = "bold", color = "black") +
  scale_fill_manual(values = baseball_colors) +
  labs(
    title = "Shohei Ohtani's Weighted Score 2024 vs. League",
    subtitle = "Comparison with Zack Wheeler and Total Mean",
    x = "Weighted Score 2024",
    y = "Category"
  ) +
  theme_minimal(base_size = 14)
```

Cleveland Dot Plot that helps compare where Ohtani places against other pitchers
```{r}
top_15 <- head(final_stats_extended[order(-final_stats_extended$Weighted_Score2024), ], 15)

ggplot(top_15, aes(x = Weighted_Score2024, y = reorder(Player, Weighted_Score2024))) +
  geom_point(aes(color = ifelse(Player == "Shohei Ohtani", "Ohtani", "Others")), size = 4, show.legend = FALSE) + 
  scale_color_manual(values = c("Ohtani" = "dodgerblue", "Others" = "gray50")) +
  labs(
    title = "Cleveland Dot Plot of Top 15 Players",
    x = "Weighted Score 2024",
    y = "Players"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5)
  )
```

